---
title: "rr_tsa_shinyapp"
author: "Mateusz Kijewski"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rr_tsa_shinyapp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## General description

The project enables user to run Shiny app for Time series analysis. Also it implements
S4 class tsExt which extends the xts class with more methods for plotting or testing the time series.


## User step by step guide

The first page of the application is the Home page – here a user can upload a .CSV file which
will be used in further steps. The file needs to be the time series format (first column is data index
and second one is time series) – we provide a sample dataset to test the application (demo.csv).
There is also a “Create RMarkdown report” button on this page, which can be used to generate a
file with the analysis with default parameters.

In the second tab (Overview) there are presented some information about the time series –
at the top of the page there is a plot which presents the time series, and at the bottom – basic
statistics like minimum and maximum value, mean, frequency of the data and range of selected days.
The title of the plot, as well as frequency and range of the data, can be set at the sidebar on the left.

The next two pages – Seasonality and Stationarity – are the tabs that enable to decompose
the time series. A user can choose a seasonal component, select test to perform (Kruskal-Wallis, QS,
KPSS test, Augmented Dickey-Fuller) or differentiate the time series the desired number of times.
The last tab is a tool for forecasting – here a user can choose a number of observations for
the testing sample used in prediction, and the method of estimation: Naïve, ARIMA, or EMA. 

## S4 Class description

The tsExt class has following methods:

``` {}
# Setting values for class attributes
f = "set_values", signature = c(object="tsExt", dates="ANY", freq="ANY", name="ANY")

# Summarize the time series
f = "summary", signature = "tsExt"

# Plot the time series using dygraphs
f = "plot_ts", signature = c("tsExt")

# Plot decomposition of time series or differenced time series depending on type argument
f = "plot_test", signature = c(object="tsExt", type="ANY", n_dif="ANY", freq="ANY", seas_type="ANY")

# Plot decomposition of time series or differenced time series depending on type argument
f = "plot_test", signature = c(object="tsExt", type="ANY", n_dif="ANY", freq="ANY", seas_type="ANY")

# Plot decomposition of time series or differenced time series depending on type argument
f = "plot_test", signature = c(object="tsExt", type="ANY", n_dif="ANY", freq="ANY", seas_type="ANY")

# Tests the stationarity of time series
f = "test_station", signature = c("tsExt", "character", "numeric")

# Tests the seasonality of time series
f = "test_season", signature = c("tsExt", "character")
```

tsExt instance example:
```{r, echo=FALSE, message=F, warning=F}
library("shiny")
library("shinyjs")
library("shinythemes")
library("shinydashboard")
library("tidyverse")
library("quantmod")
library('dygraphs')
library('dplyr')
library('lubridate')
library('seasonal')
library('tseries')
library('forecast')
library('assertthat')
## defining class
check_input <- function(object) {
  errors <- character()
  if (nrow(object@ts) == 0) {
    msg <- "No valid data provided, 0 complete cases"
    errors <- c(errors, msg)
  }
  if (!(object@freq %in% c('daily', 'weekly', 'monthly', 'quarterly', 'yearly'))) {
    msg <- "Wrong data frequency, choose from: 'daily', 'weekly', 'monthly', 'quarterly', 'yearly' "
    errors <- c(errors, msg)
  }
  
  if (length(errors) == 0) TRUE else errors
}
tsExt <- setClass('tsExt',
                  slots = c(ts = "xts",
                            name = "character",
                            freq = "character",
                            base_ts = "xts"),
                  prototype = c(name = "", freq="daily", base_ts=xts()),
                  validity = check_input
)

#init
setMethod(f = "initialize", signature = "tsExt",
          function( .Object, ts, name, freq) 
          { 
            .Object@ts <- ts
            .Object@name <- name
            .Object@freq <- freq
            if(length(names(.Object@ts)) != 1)
              warning('Wrong input format, selected only first column')
            .Object@ts <- .Object@ts[,1]
            .Object@ts[,1] <- as.numeric(.Object@ts[,1])
            .Object@ts <- .Object@ts[complete.cases(.Object@ts ),]
            #create copy of base ts for further analysis
            .Object@base_ts <- .Object@ts
            
            .Object@ts <- get(paste0('apply.',.Object@freq))(.Object@base_ts, FUN=mean)
            return(.Object)
          } 
)

#set values
setGeneric(name = "set_values",function(object, dates, freq, name) standardGeneric("set_values"))
setMethod(f = "set_values", signature = c(object="tsExt", dates="ANY", freq="ANY", name="ANY"), 
          definition = function(object, dates, freq, name){
            if(!missing(freq)) object@freq <- freq
            if(!missing(name)) object@name <- name
            if(!missing(dates)) object@ts <- get(paste0('apply.',object@freq))(object@base_ts[dates], FUN=mean)
            else object@ts <- get(paste0('apply.',object@freq))(object@base_ts, FUN=mean)
            return(object)
          })
#new summary
setMethod(f = "summary", signature = "tsExt",
          function(object) 
          { 
            data.frame(
              name = object@name,
              min = min(object@ts),
              max = max(object@ts),
              mean = mean(object@ts),
              freq = object@freq,
              days = as.numeric(max(index(object@ts)) - min(index(object@ts)))
              
            )
          } 
)

#plot
# plot time series
setGeneric(name = "plot_ts",function(object) standardGeneric("plot_ts"))
setMethod(f = "plot_ts", signature = c("tsExt"),
          function(object) 
          { 
            dygraph(object@ts, main=object@name, width = (1000), height = (500)) %>%
              dyAxis("x", label = "Date", axisLabelFormatter = "function(d) { return d.getFullYear() }", 
                     ticker="function(a, b, pixels, opts, dygraph, vals) {
                     return Dygraph.getDateAxis(a, b, Dygraph.ANNUAL, opts, dygraph)}") %>%
              dyAxis("y", label = "Values")
            } 
)

# plot stationarity/seasonality
setGeneric(name = "plot_test",function(object, type, n_dif, freq, seas_type) standardGeneric("plot_test"))

setMethod(f = "plot_test", signature = c(object="tsExt", type="ANY", n_dif="ANY", freq="ANY", seas_type="ANY"),
          definition = function(object, type, n_dif, freq, seas_type){
            if(missing(type)) type <- 'station'
            if(type == 'station'){
              if(missing(n_dif)) n_dif <- 1
              validate_that(n_dif%%1==0,msg="n_dif must be an integer!")
              validate_that(nrow(object@ts)>0,msg="Number of rows must be higher than number of diff")
              data <- object@ts
              print(data[1:5])
              if(n_dif>=1){
                for( i in 1:n_dif)
                  data <- diff(data)
              }
              dygraph(data, main=paste(object@name, 'diff:',n_dif), width = (1000), height = (500)) %>%
                dyAxis("x", label = "Date", axisLabelFormatter = "function(d) { return d.getFullYear() }", 
                       ticker="function(a, b, pixels, opts, dygraph, vals) {
                       return Dygraph.getDateAxis(a, b, Dygraph.ANNUAL, opts, dygraph)}") %>%
                dyAxis("y", label = "Values")
              
          } else if(type == 'season'){
            if(missing(seas_type)) seas_type <- 'additive'
            if(missing(freq)) freq <- object@freq
            freq_tab <-  c(365,52,12,4,2)
            freq_names <- c('daily', 'weekly', 'monthly', 'quarterly', 'yearly')
            validate_that(freq %in% freq_names,
                          msg="Wrong data frequency, choose from: 'daily', 'weekly', 'monthly', 'quarterly', 'yearly'")
            validate_that(seas_type %in% c('multiplicative', 'additive'), msg="Wrong seasonality type, choose from: 'multiplicative', 'additive'")
            validate_that(match(freq,freq_names)>=match(object@freq,freq_names), msg="Wrong seasonality type, choose from: 'multiplicative', 'additive'")
            plot(decompose(ts(object@ts, frequency = freq_tab[match(object@freq,freq_names)]), type=seas_type))
          } else {
            stop("Wrong type, choose from: 'station', 'season'")
          }
            
          }
)



# provide statistics for stationarity/seasonality

setGeneric(name = "test_station",function(object, test, n_dif) standardGeneric("test_station"))

setMethod(f = "test_station", signature = c("tsExt", "character", "numeric"),
          definition = function(object, test, n_dif){
            data <- object@ts
            if(n_dif>=1){
              for( i in 1:n_dif)
                data <- diff(data)
            }
            data <- data[(n_dif+1):nrow(data)]
            print(data[1:5])
            if(test == 'all'){
              a_test <- adf.test(data)
              k_test <- kpss.test(data)
              data.frame(
                name = c("ADF test","KPSS test"),
                null = c("Non-stationary data", "Stationary data"),
                stat_val = c(as.numeric(a_test$statistic),as.numeric(k_test$statistic)),
                pval = c(as.numeric(a_test$p.value),as.numeric(k_test$p.value))
              )
            } else if(test == 'kpss'){
              k_test <- kpss.test(data)
              data.frame(
                name = c("KPSS test"),
                null = c("Non-stationary data"),
                stat_val = c(as.numeric(k_test$statistic)),
                pval = c(as.numeric(k_test$p.value))
              )
            } else if(test == 'adf'){
              a_test <- adf.test(data)
              data.frame(
                name = c("ADF test"),
                null = c("Non-stationary data"),
                stat_val = c(as.numeric(a_test$statistic)),
                pval = c(as.numeric(a_test$p.value))
              )
            }  else {
              stop("Wrong test, choose from: 'all', 'adf', 'kpss")
            }
          }
)

# provide statistics for stationarity/seasonality

# provide statistics for stationarity/seasonality

setGeneric(name = "test_season",function(object, test) standardGeneric("test_season"))

setMethod(f = "test_season", signature = c("tsExt", "character"),
          definition = function(object, test){
            a <- as.numeric(object@ts)
            freq_tab <-  c(365,52,12,4,2)
            freq_names <- c('daily', 'weekly', 'monthly', 'quarterly', 'yearly')
            f <- freq_tab[match(object@freq, freq_names)]
            b <- rep(1:f,length(a))[1:length(a)]
            
            if(test == 'all'){
              qs <- qs(seas(ts(apply.monthly(object@ts,FUN=mean),start=1950,frequency=12)))
              c <- kruskal.test(a ~ b)
              data.frame(
                name = c("Kruskal-Wallis test","QS test"),
                null = c("Seasonal component don't exists","Seasonal component don't exists"),
                stat_val = c(as.numeric(c$statistic),as.numeric(qs[,1][1])),
                pval = c(as.numeric(c$p.value),as.numeric(qs[,2][1]))
              )
            } else if(test == 'kw'){
              c <- kruskal.test(a ~ b)
              data.frame(
                name = "Kruskal-Wallis test",
                null = "Seasonal component don't exists",
                stat_val = as.numeric(c$statistic),
                pval = as.numeric(c$p.value)
              )
            } else if(test == 'qs'){
              qs <- qs(seas(ts(apply.monthly(object@ts,FUN=mean),start=1950,frequency=12)))
              data.frame(
                name = "QS test",
                null = "Seasonal component don't exists",
                stat_val = as.numeric(qs[,1][1]),
                pval = as.numeric(qs[,2][1])
              )
            } else {
              stop("Wrong type, choose from: 'all', 'kw', 'qs'")
            }
          }
)
# predict values

setGeneric(name = "predict_ts",function(object, steps, method, graph) standardGeneric("predict_ts"))

setMethod(f = "predict_ts", signature = c("tsExt", "numeric", "character", "logical"),
          definition = function(object, steps, method, graph){
            data = object@ts
            split = nrow(object@ts)-steps
            training = as.numeric(object@ts[1:split])
            testing = as.numeric(object@ts[(split+1):nrow(object@ts)])
            if(method == 'all'){
              pred1 = as.numeric(ses(training,steps)$mean)
              data$ema <- c(training,pred1)
              err1 = testing - pred1
              mod = auto.arima(training)
              pred2 = as.numeric(forecast(mod,h=steps)$mean)
              data$arima <- c(training,pred2)
              err2 = testing - pred2 
              pred3 = as.numeric(naive(training,steps)$mean)
              data$naive <- c(training,pred3)
              err3 = testing - pred3
              if(graph){
                return({
                  dygraph(data[(nrow(data)-2*steps):nrow(data)], main=paste("Comparison of three methods for",object@name), width = (1000), height = (500)) %>%
                    dyAxis("x", label = "Date", axisLabelFormatter = "function(d) { return d.getFullYear() }", 
                           ticker="function(a, b, pixels, opts, dygraph, vals) {
                           return Dygraph.getDateAxis(a, b, Dygraph.ANNUAL, opts, dygraph)}") %>%
                    dyAxis("y", label = "Values")
              })
              }
              statistics <- data.frame()
              for(j in list(err1,err2,err3)){
                i <- j[1]
                MSE <- mean(abs(i)^2)
                RMSE <- sqrt(MSE)
                MAE <- mean(abs(i))
                MAPE <- mean(abs(i)/testing)
                next_row <- data.frame(MSE,RMSE,MAE,MAPE)
                statistics <- rbind(statistics,next_row)
              }
              statistics
              } else if(method == 'ema'){
                pred = as.numeric(ses(training,steps)$mean)
                err = testing - pred
                data$ema <- c(training,pred)
                if(graph){
                  return({
                    dygraph(data[(nrow(data)-2*steps):nrow(data)], main=paste("Exponential smoothing for ",object@name), width = (1000), height = (500)) %>%
                      dyAxis("x", label = "Date", axisLabelFormatter = "function(d) { return d.getFullYear() }", 
                             ticker="function(a, b, pixels, opts, dygraph, vals) {
                             return Dygraph.getDateAxis(a, b, Dygraph.ANNUAL, opts, dygraph)}") %>%
                      dyAxis("y", label = "Values")
                })
                }
                
                statistics <- data.frame()
                for(j in list(err)){
                  i <- j[1]
                  Name <- 'EMA'
                  MSE <- mean(abs(i)^2)
                  RMSE <- sqrt(MSE)
                  MAE <- mean(abs(i))
                  MAPE <- mean(abs(i)/testing)
                  next_row <- data.frame(MSE,RMSE,MAE,MAPE)
                  statistics <- rbind(statistics,next_row)
                }
                statistics
                } else if(method == 'arima'){
                  mod = auto.arima(training)
                  pred = as.numeric(forecast(mod,h=steps)$mean)
                  err = testing - pred
                  data$arima <- c(training,pred)
                  if(graph){
                    return({
                      dygraph(data[(nrow(data)-2*steps):nrow(data)], main=paste("Arima prediction for ",object@name), width = (1000), height = (500)) %>%
                        dyAxis("x", label = "Date", axisLabelFormatter = "function(d) { return d.getFullYear() }", 
                               ticker="function(a, b, pixels, opts, dygraph, vals) {
                               return Dygraph.getDateAxis(a, b, Dygraph.ANNUAL, opts, dygraph)}") %>%
                        dyAxis("y", label = "Values")
                  })
                  }
                  statistics <- data.frame()
                  for(j in list(err)){
                    i <- j[1]
                    Name <- 'ARIMA'
                    MSE <- mean(abs(i)^2)
                    RMSE <- sqrt(MSE)
                    MAE <- mean(abs(i))
                    MAPE <- mean(abs(i)/testing)
                    next_row <- data.frame(MSE,RMSE,MAE,MAPE)
                    statistics <- rbind(statistics,next_row)
                  }
                  statistics
                  } else if(method == 'naive'){
                    pred = as.numeric(naive(training,steps)$mean)
                    err = testing - pred
                    data$naive <- c(training,pred)
                    if(graph){
                      return({
                        dygraph(data[(nrow(data)-2*steps):nrow(data)], main=paste("Naive prediction for ",object@name), width = (1000), height = (500)) %>%
                          dyAxis("x", label = "Date", axisLabelFormatter = "function(d) { return d.getFullYear() }", 
                                 ticker="function(a, b, pixels, opts, dygraph, vals) {
                                 return Dygraph.getDateAxis(a, b, Dygraph.ANNUAL, opts, dygraph)}") %>%
                          dyAxis("y", label = "Values")
                    })
                    }
                    statistics <- data.frame()
                    for(j in list(err)){
                      i <- j[1]
                      Name <- 'Naive'
                      MSE <- mean(abs(i)^2)
                      RMSE <- sqrt(MSE)
                      MAE <- mean(abs(i))
                      MAPE <- mean(abs(i)/testing)
                      next_row <- data.frame(MSE,RMSE,MAE,MAPE)
                      statistics <- rbind(statistics,next_row)
                    }
                    statistics
                    } else {
                      stop("Wrong type, choose from: 'arima', 'ema', 'naive'")
                    }
            
                  }
)
```


``` {r}
test = read.table('../sample_data/demo.csv', header=TRUE, sep=',', colClasses=c("Date", rep("numeric", 6)))
data = as.xts(test[-1], order.by=test[[1]])
data = tsExt(data, name='Test', 'daily')

plot_ts(data)

summary(data)

test_season(data, 'all')

plot_test(data, type = 'season', seas_type='multiplicative')

test_station(data, 'all', 1)

plot_test(data, type = 'station', n_dif=1)

```

